---
title: "Forensics road crashes starter"
author: "Leah Kemp"
date: "3/2/2021"
output:
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)
```

# Introduction

This is an example for how to get started with analysing a road crash dataset (on a dummy/fake dataset). I'll be using predominantly [dplyr](https://dplyr.tidyverse.org/) since it's an intuitive package for carrying out data manipulations that I really like to use!

*Note: I reduced the size of the dataset for speed*

# Load libraries

```{r}
library(xlsx)
library(dplyr)
```

# Load data

```{r, error=TRUE}
# Reads sheet 1 of the spreadsheet
data <- read.xlsx("/home/lkemp/forensics_road_crashes_starter/dummy.xlsx", 1, colNames = TRUE, rowNames = TRUE)
```

Throws an error because of lack of java memory, probably because it a large file?

Solution found on [stack overflow](https://stackoverflow.com/questions/21937640/handling-java-lang-outofmemoryerror-when-writing-to-excel-from-r)

```{r, error=TRUE}
options(java.parameters = "-Xmx1000m")
data <- read.xlsx("/home/lkemp/forensics_road_crashes_starter/dummy.xlsx", 1, colNames = TRUE, rowNames = TRUE)
```

Still getting an error...found another solution on [stack overflow](https://stackoverflow.com/questions/19147884/importing-a-big-xlsx-file-into-r) using an additional library to load the data

```{r}
library(openxlsx)
data <- read.xlsx("/home/lkemp/forensics_road_crashes_starter/dummy.xlsx", 1, colNames = TRUE, rowNames = TRUE)
```

Great that loaded the data!

However, the `Date.of.crash` column isn't formatted quite right:

```{r}
head(data)
```

We might be able to deal with this when loading the data using the `detectDates = TRUE` argument

```{r}
data <- read.xlsx("/home/lkemp/forensics_road_crashes_starter/dummy.xlsx", 1, colNames = TRUE, rowNames = TRUE, detectDates = TRUE)
```

See if it worked:

```{r}
head(data)
```

Looks like it did!

*Note. you might need to convert the "x" cell values to "NA" for handeling them in R*

# Data cleaning

Convert cells with `x` to `NA` for better handling in R

```{r}
data <- data %>%
  mutate(across(where(is.character), ~na_if(., "x")))
```

Check it worked ok

```{r}
data %>%
  head()
```

Looks good!

Re-name the column names with `#` in them (R reads this as a comments and won't evaluate the code)

```{r}
data <- data %>%
  rename(no.deceased = "#.deceased")

data <- data %>%
  rename(no.vehicles.involved = "#.vehicles.involved")
```

Now I want to change some of the variables

```{r}
data$speed <- as.numeric(data$speed)
data$no.deceased <- as.numeric(data$no.deceased)
data$year <- as.character(data$year)
```

# Summarise the data

## Total number of drivers

Get the number of rows in the dataframe, and therefore the total number of drivers (assuming each row represents a distinct driver)

```{r}
data %>%
  count()
```

99 individuals were involved in road crashes (note. I reduced the dataset so this is less)

## Number of drivers over a certain age

This time we do the same thing (counting the number of rows) but after we have "filtered" (removed the rows) for which `Age` is greater than 60

```{r}
data %>%
  filter(Age > 60) %>%
  count()
```

20 of these 99 people involved in road crashes were over 60 years old 

Another thing of interest, "cutting" up a variable (in this case "Age") into portions and summarising over it

```{r}
data %>%
  group_by(Age.range = cut(Age, breaks= seq(0, 120, by = 10))) %>%
  summarise(num_individuals = n()) %>%
  arrange(as.numeric(Age.range))
```

Try the same but looking at a different variable (Blood.alcohol) and different age ranges

```{r}
data %>%
  group_by(Age.range = cut(Age, breaks= seq(0, 120, by = 20))) %>%
  summarise(mean_blood_alcohol = mean(Blood.alcohol)) %>%
  arrange(as.numeric(Age.range))
```

## Mean and standard deviations

```{r}
data %>%
  # Choose which variables to look at (you can use more or less variables)
  group_by(District, Gender) %>%
  # Choose the variable to summarise (in this case I used "Age")
  summarise(mean_age = mean(Age), sd_age = sd(Age))
```

```{r}
data %>%
  # Try looking at only "Day.of.Week" 
  group_by(Day.of.Week) %>%
  # Try looking at a different variable (in this case I used "speed")
  summarise(mean_Blood.alcohol = mean(Blood.alcohol), sd_Blood.alcohol = sd(Blood.alcohol))
```

These results can be written to file:

```{r}
data %>%
  group_by(District) %>%
  summarise(mean_speed = mean(speed), sd_speed = sd(speed)) %>%
  write.csv("speed_by_district.csv", row.names = FALSE)
```

## Max and min

```{r}
data %>%
  group_by(District, Gender) %>%
  summarise(min_blood_alcohol = min(Blood.alcohol), max_blood_alcohol = max(Blood.alcohol))
```

## Number of deceased drivers with alcohol in their blood

Create a new column/variable that is "yes" or "no" to presence of alcohol in blood based on the `Blood.alcohol` column

```{r}
data <- data %>%
  mutate(alcohol_in_system = case_when(
    Blood.alcohol > 0 ~ "yes",
    Blood.alcohol == 0 ~ "no"))
```

Get the number of rows in the dataframe, and therefore the total number of drivers (assuming each row represents a distinct driver) after we have "filtered" (removed the rows) for which `alcohol_in_system` is yes

```{r}
data %>%
  filter(alcohol_in_system == "yes") %>%
  count()
```

An alternative way of doing this, filter for rows/cases where at least one person was deceased and get the number of rows/cases where there was/wasn't any alcohol in their system

```{r}
data %>%
  filter(no.deceased > 0) %>%
  group_by(alcohol_in_system) %>%
  summarise(count = n())
```

# Useful links

- [A good starter with dplyr](https://dplyr.tidyverse.org/)
- [Compute summary statistics in R](https://www.datanovia.com/en/lessons/compute-summary-statistics-in-r/)
- [Aggregating and analyzing data with dplyr](https://datacarpentry.org/R-genomics/04-dplyr.html)


















